---
title: "Deliverable II"
author: "Oeuyown Kim"
date: "February 16, 2022"
output: html_document
---

This analysis uses data from the Race and Social Equity (RSE) Index from the City of Seattle and Covid-19 case data from King County to explore the latent concept of being disadvantaged. The disadvantage measure will be used to analyze how tracts with higher disadvantage scores were impacted by Covid-19.

# Cluster Analysis

## Prepare Data

```{r libraries, message=FALSE}
# load libraries
library(tidyverse)
library(dplyr)
```

**Read in data from the Race and Social Equity Index from the City of Seattle:**
```{r prepRSE}
link='https://raw.githubusercontent.com/oeuyown/computational-thinking/main/deliverable_1/data/Racial_and_Social_Equity_Composite_Index.csv'

# read in data
rseIndex = read.csv(link)

# reset indexes to R format:
row.names(rseIndex)=NULL

# omit NA values
rseIndex = na.omit(rseIndex)

# sort columns by GEOID
rseIndex = arrange(rseIndex, GEOID10)
```

### I. Data to Cluster

**a. Subset the data **

```{r subsetting}
selection=c("GEOID10","COMPOSITE_PERCENTILE", "RACE_ELL_ORIGINS_PERCENTILE","SOCIOECONOMIC_PERCENTILE","HEALTH_PERCENTILE")

dataToCluster=rseIndex[,selection]
```

**b. Set labels as row index **

```{r rowNames, eval=TRUE}
dataToCluster$GEOID10=NULL
row.names(dataToCluster)=dataToCluster$GEOID10
```


**c. Decide if data needs to be transformed  **

The data does not need to be transformed because they all fall within a similar range (they are on the same scale) and there are no outliers.

```{r boxPlotS, eval=TRUE}
boxplot(dataToCluster,horizontal = T, las=2,cex.axis=0.4)
```

### II. Compute the DISTANCE MATRIX:

**d. Set random seed **

```{r clusterSeed, eval=TRUE}
set.seed(999)
```

**e. Decide distance method and compute distance matrix **

```{r cluster_DistanceMatrix, eval=TRUE, message=FALSE}
library(cluster)
dataToCluster_DM=daisy(x=dataToCluster, metric = "gower")
```

## Compute Clusters

### 1. Apply function

**a. Estimate optimal number of clusters**
The optimal number of clusters is 4.

```{r gapStat, message=FALSE}
library(factoextra)
gap_stat = clusGap(dataToCluster, FUN = kmeans, K.max = 24, B = 50)

fviz_gap_stat(gap_stat) + theme_minimal() + ggtitle("Gap Statistic")
```

```{r clusterComputeALL, eval=TRUE}

NumberOfClusterDesired=4

# Partitioning technique
res.pam = pam(x=dataToCluster_DM,
              k = NumberOfClusterDesired,
              cluster.only = F)

# Hierarchical technique- agglomerative approach
library(factoextra)
res.agnes= hcut(dataToCluster_DM, 
                k = NumberOfClusterDesired,
                isdiss=TRUE,
                hc_func='agnes',
                hc_method = "ward.D2")

# Hierarchical technique- divisive approach
res.diana= hcut(dataToCluster_DM, 
                k = NumberOfClusterDesired,
                isdiss=TRUE,
                hc_func='diana',
                hc_method = "ward.D2")
```

### 2. Clustering results. 

**2.1 Add results to original data frame**

```{r clusterSave_toDF, eval=TRUE}
rseIndex$pam=as.factor(res.pam$clustering)
rseIndex$agn=as.factor(res.agnes$cluster)
rseIndex$dia=as.factor(res.diana$cluster)
```

**2.2 Verify ordinality in clusters**

Pam Check
```{r clusPamCheck, echo=FALSE}
aggregate(data=rseIndex,
          COMPOSITE_PERCENTILE~pam,
          FUN=mean)
```

Agnes Check
```{r clusAgnCheck, echo=FALSE}
aggregate(data=rseIndex,
          COMPOSITE_PERCENTILE~agn,
          FUN=mean)
```

Diana Check
```{r clusDiaCheck, echo=FALSE}
aggregate(data=rseIndex,
          COMPOSITE_PERCENTILE~dia,
          FUN=mean)
```

### 3. Evaluate Results.

**3.1 Plot silhouettes**

Pam Silhouette
```{r clust_silhou_PAM, echo=FALSE}
fviz_silhouette(res.pam)
```

Agnes Silhouette
```{r clust_silhou_AGNES, echo=FALSE}
fviz_silhouette(res.agnes)
```

Diana Silhouette:
This is the best clustering method for this set of data because it's the only one without negative values and it does not return any observations that are badly clustered, unlike the Pam and Agnes methods.

```{r clust_silhou_DIANA, echo=FALSE}
library(factoextra)
library(ggplot2)
fviz_silhouette(res.diana)
```

**3.2 Detecting cases badly clustered**

```{r negativeSILs, eval=TRUE}
pamEval=data.frame(res.pam$silinfo$widths)
agnEval=data.frame(res.agnes$silinfo$widths)
diaEval=data.frame(res.diana$silinfo$widths)

pamPoor=rownames(pamEval[pamEval$sil_width<0,])
agnPoor=rownames(agnEval[agnEval$sil_width<0,])
diaPoor=rownames(diaEval[diaEval$sil_width<0,])

diaPoor
```

## Compare Clustering

**Prepare a bidimensional map using distances**

```{r cmd_Map, eval=TRUE}
projectedData = cmdscale(dataToCluster_DM, k=2)

# save coordinates to original data frame:
rseIndex$dim1 = projectedData[,1]
rseIndex$dim2 = projectedData[,2]
```

**See the "map"**

```{r plotCmdmap, eval=TRUE}
base= ggplot(data=rseIndex,
             aes(x=dim1, y=dim2,
                 label=GEOID10)) 
base + geom_text(size=2)
```

**Plot results from PAM**  
```{r plotpam, echo=FALSE}
pamPlot=base + labs(title = "PAM") + geom_point(size=2,
                                              aes(color=pam),
                                              show.legend = T)  
pamPlot
```

**Plot results from Hierarchical AGNES**  
```{r plotagn, echo=FALSE}
agnPlot=base + labs(title = "AGNES") + geom_point(size=2,
                                              aes(color=agn),
                                              show.legend = T) 
agnPlot
```

**Plot results from Hierarchical DIANA**  
```{r plotdia, echo=FALSE}
diaPlot=base + labs(title = "DIANA") + geom_point(size=2,
                                              aes(color=dia),
                                              show.legend = T) 
diaPlot
```

**Compare visually**

```{r plotcompare, echo=FALSE, message=FALSE}
library(ggpubr)

ggarrange(pamPlot, agnPlot, diaPlot,ncol = 3,common.legend = T)
```

**Annotating outliers**

```{r plotpam2, echo=FALSE, message=FALSE}
library(ggrepel)

LABELpam=ifelse(rseIndex$GEOID10%in%pamPoor,rseIndex$GEOID10,"")
LABELdia=ifelse(rseIndex$GEOID10%in%diaPoor,rseIndex$GEOID10,"")
LABELagn=ifelse(rseIndex$GEOID10%in%agnPoor,rseIndex$GEOID10,"")

pamPlot + geom_text_repel(aes(label=LABELpam))
```

```{r plotdia2, echo=FALSE}
diaPlot + geom_text_repel(aes(label=LABELdia))
```

```{r plotagn2, echo=FALSE}
agnPlot + geom_text_repel(aes(label=LABELagn))
```

**The Dendogram (for hierarchical approaches)**

```{r clusagnREPORTdendo, echo=FALSE, message=FALSE}
fviz_dend(res.agnes,k=NumberOfClusterDesired, cex = 0.45, horiz = T,main = "AGNES approach")
```

```{r clusdiaREPORTdendo, echo=FALSE}
fviz_dend(res.diana,k=NumberOfClusterDesired, cex = 0.45, horiz = T,main = "DIANA approach")
```

## Factor Analysis

### Subset the Origina Data Frame

```{r subsetFA, eval=TRUE}
selection=c("GEOID10","COMPOSITE_PERCENTILE", "RACE_ELL_ORIGINS_PERCENTILE","SOCIOECONOMIC_PERCENTILE","HEALTH_PERCENTILE")

dataForFA=rseIndex[,selection]
```

### Confirmatory Factor Analysis

```{r checkNames, eval=TRUE}
names(dataForFA)
```

```{r predictFit, eval=TRUE, message=FALSE}
library(lavaan)

model='equity=~RACE_ELL_ORIGINS_PERCENTILE + SOCIOECONOMIC_PERCENTILE + HEALTH_PERCENTILE'
  
fit<-cfa(model, data = dataForFA,std.lv=TRUE)
indexCFA=lavPredict(fit)
```

```{r indexCFA, eval=TRUE, message=FALSE}
indexCFA[1:10]
```

**Force the data to return to "0 to 10"**

```{r indexNorm, eval=TRUE, message=FALSE}
library(scales)
indexCFANorm=rescale(as.vector(indexCFA), 
                     to = c(0, 10))
indexCFANorm[1:10]
```

```{r indexCFANorm, eval=TRUE}
rseIndex$demo_FA=indexCFANorm
rseIndex$demo_FA
```

**Compare the new index with the original score**

```{r plotIndex, eval=TRUE}
base=ggplot(data=rseIndex,
            aes(x=demo_FA,y=COMPOSITE_PERCENTILE))
base+geom_point()
```

**Evaluation measures of our index for equity**

```{r loadings, eval=TRUE}
evalCFA1=parameterEstimates(fit, standardized =TRUE)

evalCFA1[evalCFA1$op=="=~",c('rhs','std.all','pvalue')]
```

```{r coeff, eval=TRUE}
evalCFA2=as.list(fitMeasures(fit))
```

Chi-Squared

```{r chisq, eval=TRUE}
evalCFA2[c("chisq", "df", "pvalue")] 
```

Tucker-Lewis

```{r, eval=TRUE}
evalCFA2$tli # > 0.90
```

RMSEA

```{r RMSEA, eval=TRUE}
evalCFA2[c( 'rmsea.ci.lower','rmsea','rmsea.ci.upper')] 
```

**SEM Model**

```{r semPlot, eval=TRUE, message=FALSE}
library(semPlot)
semPaths(fit, what='std', nCharNodes=0, sizeMan=12,
         edge.label.cex=1.5, fade=T,residuals = F)
```

# Basic Modeling in R

## Prepare Data

Read in data for COVID cases by census tract from King County:
```{r prepCovid, echo=FALSE}
covidLink="https://raw.githubusercontent.com/oeuyown/computational-thinking/main/deliverable_2/covidcases_bytract.csv"

# read in data
covidData=read.csv(covidLink)

# view data
covidData %>%
  head()
```

Check column names for `covidData`
```{r covidCols}
covidData %>% 
  colnames()
```

Merge `rseIndex` data with `covidData` by GEOID
```{r mergeData, eval=TRUE}
# merge datasets
demoCovid=merge(rseIndex,covidData,by.x = "GEOID10",by.y = "ï..Location_Name")

# omit NA values
demoCovid = na.omit(demoCovid)

```

Verifying data structure:

```{r, eval=TRUE}
str(demoCovid,width = 70,strict.width='cut')
```

## Continuous outcome 

### Explanatory Approach
The explanatory approach was chosen because the Rsquared from _postResample_ from the predictive approach is below 0.5.

1. State hypotheses:

```{r, eval=TRUE}
# hypothesis 1: The disadvantage score increases as Confirmed Covid Case Rates increase:
hypo1=formula(COMPOSITE_PERCENTILE ~ Confirmed_Case_Rate)

# hypothesis 2: The disadvantage score increases as Confirmed Covid Case Rates, Hospitalization Rates, and Death Rates increase:

hypo2=formula(COMPOSITE_PERCENTILE~ Confirmed_Case_Rate + Hospitalization_Rate + Death_Rate)
```

2. Compute regression models:

```{r, eval=TRUE}
# results
gauss1=glm(hypo1,
           data = demoCovid,
           family = 'gaussian')

gauss2=glm(hypo2,
           data = demoCovid,
           family = 'gaussian')
```

3. See results:

First Hypothesis:
For every one unit increase in the `Confirmed_Case_Rate`, there is a 0.00006367 unit increase in the `COMPOSITE_PERCENTILE` (disadvantage scale), which is statistically significant.

```{r, eval=TRUE}
summary(gauss1)
```

Second Hypothesis:
Holding `Hospitalization_Rate` and `Death_Rate` constant, for every one unit increase in the `Confirmed_Case_Rate`, there is a 0.00004536 increase in the `COMPOSITIE_PERCENTILE` (disadvantage score), which is statistically significant. This rate is the only statistically significant variable. However, the `Hospitalization_Rate` is close to being statistically significant at the 10% confidence interval.
```{r, eval=TRUE}
summary(gauss2)
```

4. Search for _better_ model:

```{r, eval=TRUE}
anova(gauss1,gauss2,test="Chisq")
```

RSquared for second hypothesis

```{r, eval=TRUE, message=FALSE}
library(rsq)
rsq(gauss2,adj=T)
```

5. Verify the situation of chosen model:

5.1. The linear relationships does not hold that well:
```{r, eval=TRUE}
plot(gauss2,1)
```

5.2. Normality of residuals is assumed:

Visual exploration:
```{r, eval=TRUE}
plot(gauss2,2)
```

Mathematical exploration:
The data is normal because the p-value > 0.05
```{r, eval=TRUE}
shapiro.test(gauss2$residuals)
```

5.3. Homoscedasticity

Visual exploration:

```{r, eval=TRUE}
plot(gauss2, 3)
```

Mathematical exploration: p-value is <0.05 so I cannot assume homoscedasticity
```{r, eval=TRUE, message=FALSE}
library(lmtest)
bptest(gauss2) 
```

5.4. There is no collinearity because the VIFs are lower than 5.

```{r, eval=TRUE, message=FALSE}
library(car)
vif(gauss2) # lower than 5 is desirable
```

5.5. Analyzing the effect of atypical values

Visual exploration:
```{r, eval=TRUE}
plot(gauss2,5)
```

Querying:
```{r, eval=TRUE}
gaussInf=as.data.frame(influence.measures(gauss2)$is.inf)
gaussInf[gaussInf$cook.d,]
```


6. Summary plot

```{r, eval=TRUE, message=FALSE}
library(sjPlot)

plot_models(gauss2,vline.color = "grey")
```

